Waking up a sleeping rabbit:\\ On Natural-Language Sentence Generation with FF

### 2009-12-19 trunk/Papers/crisp-ff-10/crisp.tex
submitted version

To use TAG grammars for generation \cite{Stone2003a}, we assume a set
of ground atoms expressing the information we want the sentence to
convey, such as $\{\mathsf{sleep}(e,r_1)\}$, and a knowledge base that
contains all ground atoms we know to be true; say,
$\{\mathsf{sleep}(e,r_1), \mathsf{rabbit}(r_1), \mathsf{rabbit}(r_2),
\mathsf{white}(r_1), \mathsf{brown}(r_2)\}$.  We then add variables
ranging over constants from the knowledge base to the nodes of each
elementary tree, and equip every elementary tree with a set of atoms
over these variables to encode the meaning this tree expresses.
Fig.~\ref{fig:white-rabbit-sleeps-deriv} already shows specific
instances of the elementary trees from the grammar, in which constants
have been substituted for these variables.  The derivation in the
figure conveys the intended meaning -- in particular, that $r_1$
sleeps.  Crucially, it also describes uniquely who does the sleeping:
The sentence ``the rabbit sleeps'' would not have done this, as ``the
rabbit'' could be understood either as $r_1$ or $r_2$.  We say that
$r_2$ is a \emph{distractor} for the subject of the sentence, which is
removed by adjoining the tree for ``white''.  The sentence generation
problem consists in computing a grammatically correct TAG derivation
from instances of the elementary trees in the grammar that conveys the
intended meaning and describes all referents uniquely.

@@

More precisely, each operator has a parameter $u$ representing the
syntax node into which the elementary tree is substituted or adjoined,
and a parameter $u_i$ for each substitution node.  There are also
parameters $x_0,\ldots,x_k$ for the variables in the semantic
representation of the elementary tree; $x_0$ is the variable that
occurs at the root of the tree.  The planning state encodes a list of
possible node names; $\f{current}(u)$ expresses that $u$ is the next
node name that should be picked when a new substitution node is
created, and $\f{next}(u,v)$ says that the next node after $u$ is
$v$. These atoms are used to select names for the substitution nodes
that are introduced by adding an elementary tree; the parameter $u_n$
represents the node that will be current after the addition.

@@

The example generation problem above translates into a planning
problem whose domain is shown, in PDDL format, in
Fig.~\ref{fig:white-rabbit-as-planning}. The initial state of the
problem encodes the generation problem's knowledge base; it contains
atoms $\f{rabbit}(r_1)$, $\f{sleep}(e,r_1)$, etc.  The goal requires
syntactic completeness as $\forall u \forall A \neg \f{subst}(A,u)$
and unique reference as $\forall u \forall x \neg
\f{distractor}(u,x)$; it also specifies the semantic representation we
want to express in an atom $\f{expressed}(\mathsf{sleep},e,r_1)$.

A minimal plan for the example problem is
$\mathsf{sleeps}(\mathsf{root}, n_1, n_2, e, r_1)$;
$\mathsf{rabbit}(n_1, r_1)$; $\mathsf{white}(n_1, r_1)$. This plan can
be automatically decoded into the derivation shown in
Fig.~\ref{fig:white-rabbit-sleeps-deriv}.  The analogous plan with
$r_2$ instead of $r_1$ is not correct because the $\mathsf{sleep}$
precondition of the $\mathsf{sleeps}$ operator is not satisfied.  The
first two steps of the plan alone would not be a correct plan because
they leave an atom $\mathsf{distractor}(n_1, r_2)$ in the state, which
contradicts the goal that all distractors have been eliminated.

>>>>

To use TAG grammars for generation \cite{Stone2003a}, we assume a set
of ground atoms expressing the information we want the sentence to
convey, such as $\{\mathsf{sleep}(e,r_1)\}$, and a knowledge base that
contains all ground atoms we know to be true; say,
$\{\mathsf{sleep}(e,r_1), \mathsf{rabbit}(r_1), \mathsf{rabbit}(r_2),
\mathsf{white}(r_1), \mathsf{brown}(r_2)\}$.  We then add variables
ranging over constants from the knowledge base to the nodes of each
elementary tree, and equip every elementary tree with a set of atoms
over these variables to encode the meaning this tree expresses
(printed in italics in the figure).
Fig.~\ref{fig:white-rabbit-sleeps-deriv} already shows specific
instances of the elementary trees from the grammar, in which constants
have been substituted for these variables.  The derivation in the
figure conveys the intended meaning -- in particular, that $r_1$
sleeps.  Crucially, it also describes uniquely who does the sleeping:
The sentence ``the rabbit sleeps'' would not have done this, as ``the
rabbit'' could be understood either as $r_1$ or $r_2$.  We say that
$r_2$ is a \emph{distractor} for the subject of the sentence, which is
removed by adjoining the tree for ``white''.  In short, the sentence
generation problem consists in computing a grammatically correct TAG
derivation from instances of the elementary trees in the grammar that
conveys the intended meaning and describes all referents uniquely.

@@

More precisely, each operator has a parameter $u$ representing the
syntax node into which the elementary tree is substituted or adjoined,
and a parameter $u_i$ for each substitution node.  There are also
parameters $x_0,\ldots,x_k$ for the variables in the semantic
representation of the elementary tree; $x_0$ is the variable that
occurs at the root of the tree.  The planning state encodes a list of
possible node names: $\f{current}(u)$ expresses that $u$ is the next
node name that should be picked when a new substitution node is
created, and $\f{next}(u,v)$ says that the next node after $u$ is
$v$. These atoms are used to select names for the substitution nodes
that are introduced by adding an elementary tree; the parameter $u_n$
represents the node that will be current after the addition.

@@

The example generation problem above translates into a planning
problem whose domain is shown in
Fig.~\ref{fig:white-rabbit-as-planning}. The initial state of the
problem encodes the generation problem's knowledge base; it contains
atoms $\f{rabbit}(r_1)$, $\f{sleep}(e,r_1)$, etc.  The goal requires
syntactic completeness as $\forall u \forall A \neg \f{subst}(A,u)$
and unique reference as $\forall u \forall x \neg
\f{distractor}(u,x)$; it also specifies the semantic representation we
want to express in an atom $\f{expressed}(\mathsf{sleep},e,r_1)$.

A minimal plan for the example problem is
$\mathsf{sleeps}(\mathsf{root}, n_1, n_2, e, r_1)$;
$\mathsf{rabbit}(n_1, r_1)$; $\mathsf{white}(n_1, r_1)$. This plan can
be automatically decoded into the derivation shown in
Fig.~\ref{fig:white-rabbit-sleeps-deriv}.  The first two steps of the
plan alone would not be a correct plan because they leave an atom
$\mathsf{distractor}(n_1, r_2)$ in the state, which contradicts the
goal that all distractors have been eliminated.
